---
title: "A grammar of graphics"
author: "Michael Nelson"
date: "September 2018"
output: 
  beamer_presentation:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dev = 'pdf',
  fig.width = 4,
  fig.height = 2,
  dev.args = list(pointsize = 10),
  tidy = TRUE,
  tidy.opts = list(width = 40)
)
```

## A grammar of graphics

How do you describe a statistical graphic?

Move beyond 'named' graphics such as 'scatterplot'

A grammar?

- Defining basic components that make up a plot.
- More than a taxonomy

Based on Wilkinson (2006). Implemented in `ggplotw` by Wickham (2016)


## The components of a (statistical) graphic

- *data*: The information we want to plot (a `data.frame`)
- *geoms*: describe the type of geometric object that represent data (points, lines, bars, polygons, ...)
- *aesthetics*: the visual characteristics that represent data (position, size, shape, colour, ...)
- *coordinate system*: describes the 2-dimensional space the data is projected onto. (cartesian coordinates, polar coordinates, map projections, ...)
- *scales*: for each aesthetic: describes how the visual characteristic is converted to display values
- *stats*: describe the statistical transformations that summarize the data (counts, means, medians, regression lines, ...)
- *position*: How to adjust the position if overplotting occurs.
- *facets*: describe how the data is split into subsets and displayed as multiple smaller graphs

## A simple example

### The iris dataset

```{r example, echo = TRUE}
head(iris)
```


## A simple example

```{r simple-plot}
library(ggplot2)
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()
```

## Basic structure


 `ggplot` is the function a ggplot `object`
 
- Specify data and variables (`mapping`) in this function
- Anything specified here will be *inherited* by all layers

```{r building-blocks, eval = FALSE}
ggplot(data = iris, mapping = aes( x= Sepal.Length, y = Sepal.Width))
```

- What happens if we don't specify a layer?

## No layer

```{r ref.label='building-blocks'}
```


## A layer

```{r bb3, eval = TRUE}
ggplot(data = iris, mapping = aes( x= Sepal.Length, y = Sepal.Width)) +
 geom_point()
```

## What is a layer?

### The purpose

- To display the data - allows the viewer to see patterns, overall or local structure
- To display statistical summaries of the data


## Setting other aesthetics

### Increase the size of the points?

```{r point-size, eval = TRUE}
ggplot(data = iris, mapping = aes( x= Sepal.Length, y = Sepal.Width)) +
 geom_point(size = 3)
```

##  Mapping other aesthetics

### Differentiate the points by colour

```{r colour, eval = TRUE}
ggplot(data = iris, mapping = aes( x= Sepal.Length, y = Sepal.Width, colour = Species)) +
 geom_point()
```


##  Mapping other aesthetics

### Differentiate the points by shape

```{r colour-shape, eval = TRUE}
ggplot(data = iris, mapping = aes( x= Sepal.Length, y = Sepal.Width, colour = Species)) +
 geom_point(aes(shape = Species), size = 3)
```


## Mapping or setting aesthetics

Aesthetics can be *set* or *mapped* within a ggplot call (or individual layer).

- An aesthetic defined within `mapping = aes(...)` is mapped from the data, and a `guide` created.
- An aesthetic may also be set to a single value, by defining it outside the `mapping = aes(...)` argument